What to simulate? Inferring the direction for mental rotation

### 2014-02-01 man/cogsci-2014/cogsci-2014.tex
Tighten wording

% 
% Annual Cognitive Science Conference
% Sample LaTeX Paper -- Proceedings Format
%

@@

The only models (of which the authors are aware) that have seriously
attempted to address the decision of \textit{what} to simulate are
those by \citeA{Funt:1983wn} and \citeA{Just:1985uu}. In both of these
models, the axis and direction of rotation are computed prior to
performing the rotation. One object is then rotated through the target
rotation, and is checked against the other object for
congruency. However, this approach assumes that the corresponding
points on the two objects can be easily identified, which is not
necessarily the case. Indeed, the state-of-the-art in computer vision
suggests that there is more to shape matching than simply checking for
congruency, particularly when shapes are complex or when the shapes
might not be exactly the same
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}. Additionally, recent
research shows that when performing \textit{physical} rotations,
people do not rotate until congruency is reached, and may even rotate
\textit{away} from near perfect matches \cite{Gardony:2013gn}.

@@

The original congruency hypothesis \cite{Shepard1971} actually is a
rational solution which achieves this goal, in the sense that the
smallest amount of computation coincides with rotating through the
minimum angle. However, it violates the constraint that we do not know
points correspondences, which is exactly what necessitates the use of
imagery. Noting that a rational solution need not necessarily maintain
a single trajectory of rotation, we explore an alternative rational
model, which--rather than computing the angle of rotation ahead of
time--engages in an ``active sampling'' strategy.

@@

In the following section, we formalize this rational analysis, and
propose four models of mental rotation: one based on the original
hypothesis by \citeA{Shepard1971}; two which are extensions of the
first but with relaxed assumptions; and one based on the active
sampling approach.

@@

In the mental rotation task, people are presented with pairs of
two-dimensional shapes similar to those used by \citeA{Cooper:1975wp};
example stimuli are shown in Fig.~\ref{fig:model-traces}. The goal is
to determine whether the two images depict the same shape, or
mirror-image shapes.

@@

We define $p(\Xb\ \vert\ \Xa, \theta, h)$ to be the similarity between
$\Xb$ and a transformation of $\Xa$:
\begin{equation}
  p(\Xb\ \vert\ \Xa, \theta, h):= S(\Xb, f(\Xa, h, \theta)).
  \label{eq:likelihood}
\end{equation}
Example similarity curves are shown in Fig.~\ref{fig:model-traces}.

We do not know which vertices of $\Xb$ correspond to which vertices of
$\Xa$, so $S$ must marginalize over the set of possible mappings. For
brevity, let $\mathbf{X}_m=\M\cdot{}f(\Xa, h, \theta)$ where $\M$ is a
permutation matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{2N} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma_S^2),
  \label{eq:similarity}
\end{equation}
where $N$ is the number of vertices, $2N$ is the total number of
possible mappings,\footnote{It is $2N$ and not $N^2$ because, in polar
  coordinates, vertices are always connected to their two nearest
  neighbors in the $\theta$ dimension.}and $\sigma_S=0.15$ is the
standard deviation of the similarity.

Additionally, we assume that the observed shapes may only be
transformed by a small amount at a time, and each transformation is
costly in terms of computational resources. Specifically, if the
current mental image is $\Xt$, then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, |\epsilon|) &\mbox{ rotate by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip,} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}
where $\epsilon\sim \mathcal{N}(0, \sigma_\epsilon)$ and
$\sigma_\epsilon$ is the standard deviation of the step size in
radians.

@@

One hypothesis in the literature is that people perform a mental
rotation in the shortest direction until a match is found
\cite{Shepard1971, Cooper:1975wp}, and that determining this direction
is an implementation detail \cite{Anderson1978}, or that it is
computed beforehand \cite{Funt:1983wn}.  To reflect this hypothesis,
we created an ``oracle'' model which is told which points on each
shape correspond. The \Oc{} model first computes the correct rotation
and then simply takes actions in the shortest direction until it
reaches the point where the shapes are aligned.

@@

In the current formulation of the problem, choosing the threshold is
straightforward because we know both the exact geometry of the shapes
and that a linear transformation exists which will perfectly align
them. However, it is not always clear how to choose a reasonable
threshold \textit{a priori}, as the global optimum will depend on many
factors (such as shape complexity, dimensionality, perceptual
uncertainty, and whether the shapes are exactly identical). Thus, it
is unlikely that the \Th{} model will be robust beyond this domain.

@@

One way to deal with the problem of choosing a threshold would simply
be to perform an exhaustive search or use a global optimization
strategy. However, these strategies would not result in the linear
response time found by \citeA{Shepard1971}. A second alternative,
which could potentially demonstrate the linear response time result,
might instead just perform a Hill Climbing (\Hc{}) search for a
\textit{local} maximum. As with the \Oc{} and \Th{} models,
Eq.~\ref{eq:lh-ratio} becomes a maximum likelihood ratio. Unlike the
global maximum, however, it is not guaranteed that a local maximum of
the correct hypothesis will be greater than a local maximum of the
incorrect hypothesis.

@@

Formally, we denote $Z_h$ as our estimate of the likelihood for
hypothesis $h$, and write its distribution as:
\begin{equation}
  p(Z_h) = \int \left[\int S(\Xb, f(\Xa, \theta, h))p(\theta)\dif\theta\right] p(S)\dif S.
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use a Gaussian Process (GP) prior for $p(S)$, as in
\citeA{Osborne:2012tm}.\footnote{Because our domain is on a circle, we
  utilize a periodic kernel for the covariance
  \cite{Rasmussen:2006vz}.}  Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$,
we first place a GP prior on the log of $S_h$ in order to enforce
positivity after it is exponentiated:
\begin{equation}
\mathbb{E}[Z_h] \approx \int \exp(\mu_h(\theta))p(\theta)\dif\theta,
\end{equation}
where $\mu_h:=\mu(\log S_h)$ is the mean of the log-GP. However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP, which we denote as
$\bar{S}_h:=\exp(\mu_h)$.\footnote{Note that the approach used here
  differs slightly from the one presented in \citeA{Osborne:2012tm}
  and is based on personal communication with one of the authors
  \cite{Duvenaud:2013td}.} Then, from \citeA{Duvenaud:2013td}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

Assuming independence, we can now write $p(Z_h)\approx\mathcal{N}(Z_h\
\vert\ \mathbb{E}[Z_h], \mathbb{V}(Z_h))$, which gives us a
distribution over the likelihood ratio in Eq.~\ref{eq:lh-ratio}:
\begin{equation}
p(\ell)\approx\frac{\mathcal{N}(Z_0\ \vert\ \mathbb{E}[Z_0], \mathbb{V}(Z_0))\ p(\hi)}{\mathcal{N}(Z_1\ \vert\ \mathbb{E}[Z_1], \mathbb{V}(Z_1))\ p(\hf)}.
\end{equation}
This distribution cannot easily be calculated; however, we really are
just interested in whether $Z_0>Z_1$ or $Z_1>Z_0$. Thus, rather than
computing $p(\ell)$, we use $Z_D=Z_0-Z_1$:
\begin{equation}
  p(Z_D)\propto\mathcal{N}(\mathbb{E}[Z_0] - \mathbb{E}[Z_1], \mathbb{V}(Z_0) + \mathbb{V}(Z_1)).
\end{equation}
We then sample new observations until we are at least 95\% confident
that $Z_D\neq 0$. In other words, when $p(Z_D<0)<0.025$, we accept
$\hi$, and when $p(Z_D<0)>0.975$, we accept $\hf$. To determine which
points to sample, we can compute the expected variance of $Z_h$ if we
sampled a new observation at $\theta_a$. From \citeA{Osborne:2012tm}:
\begin{align}
\mathbb{E}&[\mathbb{V}(Z_h|\theta_a)]=\mathbb{V}(Z_h) + \mathbb{E}[Z_h] - \\
&\int \mathbb{E}[Z_h|\theta_{a}]^2 \mathcal{N}(\mu_h(\theta_a), \mathrm{Cov}_h(\theta_a, \theta_a))\dif\log S_h(\theta_a).\nonumber
\label{ref:expected-variance}
\end{align}
We compute this expected variance for each of the actions in
Eq.~\ref{eq:actions}, and then choose the action with the lowest
value.

@@

For analyses of response time, confidence intervals around harmonic
means were computed using a bootstrap analysis using 10000 bootstrap
samples (sampled with replacement). Response times were only
considered for correct responses.  We also used a bootstrap analysis
of 10000 bootstrap samples to compute the confidence intervals around
both Spearman ($r_s$) and Pearson ($r$) correlations.

For analyses of accuracy, confidence intervals were computed from a
binomial proportion with a Jeffrey's beta prior.  To test if
participants or a model was above chance on a particular stimulus, we
used the same binomial proportion, and simply tested whether $p>0.05$
for the 50\% mark.

@@

We set out to answer the question: how do people decide \textit{what}
to simulate when using mental imagery? Focusing on the specific case
of determining the direction and extent of mental rotation, we
formalized four models and compared their performance with the results
of a behavioral experiment. The \Oc{} model was the best fit to human
response times, but failed to capture anything regarding accuracy. The
\Th{} model was a slightly worse fit to human response times, and also
did not capture any trends in accuracy. The \Bq{} model fit human
reponse time data as well as the \Th{} model, and captured a small
amount of the variance in people's accuracy. The \Hc{} model barely
fit human response time data, though it did fit accuracy as well as
the \Bq{} model with an equal prior.

While it is clear that none of the models fully capture human behavior
on the mental rotation task, it is worth noting where they shed light
on aspects that have previously gone unexplained.  For example, the
\Bq{} model's adaptive stopping rule is sensitive to the prior, and
thus provides a possible explanation for why people are slower to
respond on ``flipped'' stimulus pairs.

>>>>

 
% Annual Cognitive Science Conference
% Sample LaTeX Paper -- Proceedings Format
%

@@

The only models (of which the authors are aware) that have seriously
attempted to address the decision of \textit{what} to simulate are
those by \citeA{Funt:1983wn} and \citeA{Just:1985uu}. In both of these
models, the axis and direction of rotation are computed prior to
performing the rotation. One object is then rotated through the target
rotation, and is checked against the other object for
congruency. However, this approach assumes that the corresponding
points on the two objects can be easily identified, which is not
necessarily the case.

Indeed, the state-of-the-art in computer vision suggests that there is
more to shape matching than simply checking for congruency,
particularly when shapes are complex or when the shapes might not be
exactly the same
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}. Additionally, recent
research shows that when performing \textit{physical} rotations,
people do not rotate until congruency is reached, and may even rotate
\textit{away} from near perfect matches \cite{Gardony:2013gn}.

@@

The original ``congruency'' hypothesis \cite{Shepard1971} actually is
a rational solution which achieves this goal, in the sense that the
smallest amount of computation coincides with rotating through the
minimum angle. However, it violates the constraint that we do not know
points correspondences, which is exactly what necessitates the use of
imagery. Noting that a rational solution need not necessarily maintain
a single trajectory of rotation, we explore an alternative rational
model, which--rather than computing the angle of rotation ahead of
time--engages in an ``active sampling'' strategy.

@@

@@

In this section, we formalize our rational analysis, and propose four
models of mental rotation: one based on existing models; two which are
extensions of the first but with relaxed assumptions; and one based on
the active sampling approach.

@@

We define $p(\Xb\ \vert\ \Xa, \theta, h)$ to be the similarity between
$\Xb$ and a transformation of $\Xa$: $p(\Xb\ \vert\ \Xa, \theta, h):=
S(\Xb, f(\Xa, h, \theta))$.  We do not know which vertices of $\Xb$
correspond to which vertices of $\Xa$, so $S$ must marginalize over
the set of possible mappings. For brevity, let
$\mathbf{X}_m=\M\cdot{}f(\Xa, h, \theta)$ where $\M$ is a permutation
matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{2N} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma_S^2),
  \label{eq:similarity}
\end{equation}
where $2N$ is the total number of possible mappings,\footnote{It is
  $2N$ and not $N^2$ because, in polar coordinates, vertices are
  always connected to their two nearest neighbors in the $\theta$
  dimension.}and $\sigma_S=0.15$ is the standard deviation of the
similarity. Example similarity curves are shown in
Fig.~\ref{fig:model-traces}.

Additionally, we assume that the observed shapes may only be
transformed by a small amount at a time, and each transformation is
costly in terms of computational resources. Specifically, if the
current mental image is $\Xt$, then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, \epsilon) &\mbox{ continue rotating by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip,} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}
where $\epsilon\sim \left|\mathcal{N}(0, \sigma_\epsilon)\right|$ and
$\sigma_\epsilon$ is the standard deviation of the step size in
radians.

@@

One hypothesis in the literature is that people compute the direction
and extent of rotation beforehand using \textit{a prior} knowledge of
the correspondence between points in the images
\cite{Funt:1983wn,Just:1985uu}.  To reflect this hypothesis, we
created an ``oracle'' model which is told which points on each shape
correspond. From that correspondence, it computes the correct rotation
and rotates through it.

@@

@@

In the current formulation of the problem, choosing the threshold is
straightforward because we know both the exact geometry of the shapes
and that a linear transformation exists which will perfectly align
them. However, this choice is not always clear \textit{a priori}, as
the global optimum will depend on many factors (e.g., shape
complexity, dimensionality, perceptual uncertainty, and whether the
shapes are identical).  One way to deal with the problem of choosing a
threshold would simply be to perform an exhaustive search or use a
global optimization strategy. However, these strategies would not
result in the linear response time found by \citeA{Shepard1971}. A
second alternative, which could potentially demonstrate the linear
response time result, might instead just perform a Hill Climbing
(\Hc{}) search for a \textit{local} maximum. As with the \Oc{} and
\Th{} models, Eq.~\ref{eq:lh-ratio} becomes a maximum likelihood
ratio.

@@

Formally, we denote $Z_h$ as our estimate of the likelihood for
hypothesis $h$, and write its distribution as:
\begin{equation}
  p(Z_h) = \int \left[\int S(\Xb, f(\Xa, \theta, h))p(\theta)\dif\theta\right] p(S)\dif S.
\end{equation}

This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use a Gaussian Process (GP) prior for $p(S)$, as in
\citeA{Osborne:2012tm}.  Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we
first place a GP prior on the log of $S_h$ in order to enforce
positivity after it is exponentiated, i.e. $\mathbb{E}[Z_h] \approx
\int \exp(\mu_h(\theta))p(\theta)\dif\theta$, where $\mu_h:=\mu(\log
S_h)$ is the mean of the log-GP.  However, this integral is still
intractable. To approximate it, we fit a second GP over points sampled
from the log-GP, which we denote as
$\bar{S}_h:=\exp(\mu_h)$.\footnote{Note that the approach used here
  differs slightly from the one presented in \citeA{Osborne:2012tm}
  and is based on personal communication with one of the authors
  \cite{Duvenaud:2013td}.} Then, from \citeA{Duvenaud:2013td}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

Assuming independence, we can now write $p(Z_h)\approx\mathcal{N}(Z_h\
\vert\ \mathbb{E}[Z_h], \mathbb{V}(Z_h))$, which gives us a
distribution over the likelihood ratio in Eq.~\ref{eq:lh-ratio}:
\begin{equation}
p(\ell)\approx\frac{\mathcal{N}(Z_0\ \vert\ \mathbb{E}[Z_0], \mathbb{V}(Z_0))\ p(\hi)}{\mathcal{N}(Z_1\ \vert\ \mathbb{E}[Z_1], \mathbb{V}(Z_1))\ p(\hf)}.
\end{equation}
This distribution cannot easily be calculated; however, we really are
just interested in whether $Z_0>Z_1$ or $Z_1>Z_0$. Thus, rather than
computing $p(\ell)$, we use $Z_D=Z_0-Z_1$ and compute:
$p(Z_D)\propto\mathcal{N}(\mathbb{E}[Z_0] - \mathbb{E}[Z_1],
\mathbb{V}(Z_0) + \mathbb{V}(Z_1))$.

We then sample new observations until we are at least 95\% confident
that $Z_D\neq 0$. In other words, when $p(Z_D<0)<0.025$, we accept
$\hi$, and when $p(Z_D<0)>0.975$, we accept $\hf$. To determine which
points to sample, we can compute the expected variance of $Z_h$ if we
sampled a new observation at $\theta_a$. From \citeA{Osborne:2012tm}:
\begin{align}
\mathbb{E}&[\mathbb{V}(Z_h|\theta_a)]=\mathbb{V}(Z_h) + \mathbb{E}[Z_h] - \\
&\int \mathbb{E}[Z_h|\theta_{a}]^2 \mathcal{N}(\mu_h(\theta_a), \mathrm{Cov}_h(\theta_a, \theta_a))\dif\log S_h(\theta_a).\nonumber
\label{ref:expected-variance}
\end{align}
We compute this expected variance for each of the actions in
Eq.~\ref{eq:actions}, and then choose the action with the lowest
value.

@@

For analyses of response time, confidence intervals around harmonic
means were computed using a bootstrap analysis using 10000 bootstrap
samples (sampled with replacement). Response times were only
considered for correct responses.  We also used a bootstrap analysis
of 10000 bootstrap samples to compute the confidence intervals around
both Spearman ($r_s$) and Pearson ($r$) correlations.  For analyses of
accuracy, confidence intervals were computed from a binomial
proportion with a Jeffrey's beta prior.  To test if participants or a
model was above chance on a particular stimulus, we used the same
binomial proportion, and simply tested whether $p>0.05$ for the 50\%
mark.

@@

We set out to answer the question: how do people decide \textit{what}
to simulate when using mental imagery? Focusing on the specific case
of determining the direction and extent of mental rotation, we
formalized four models and compared their performance with the results
of a behavioral experiment. While it is clear that none of the models
fully captured human behavior on the mental rotation task, it is worth
noting where they shed light on aspects that have previously gone
unexplained.  For example, the \Bq{} model's adaptive stopping rule is
sensitive to the prior, and thus provides a possible explanation for
why people are slower to respond on ``flipped'' stimulus pairs.
